---
title: '**AHS-2 Environmental Nutrition**'
output: pdf_document
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-8em} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required packages
pacs <- c("tidyverse", "readxl", "tableone", "GGally", "egg", "DescTools")
sapply(pacs, require, character.only = TRUE)

# Read data ---------------------------------------------------------------

# Data file
zipfile <- list.files(path = "./data", pattern = "\\.zip$", full.names = TRUE)
fname <- unzip(zipfile, list=TRUE)$Name[2]
# fname <- list.files(path = "./data", pattern = "\\.csv$", full.names = TRUE)

# Read data -- 88,008 obs and 190 variables
# ev <- read_csv(fname[2]) %>% 
ev <- read_csv(unz(zipfile, fname)) %>% 
  arrange(analysisid) %>% 
  mutate(edu3cat = factor(edu3cat, levels = c("Highschool", "Some College", "College Degree")),
         female = factor(female, labels = c("Male", "Female")),
         black = factor(black, labels = c("Non-Black", "Black")))

# change variable names to lower casse
names(ev) <- tolower(names(ev))

# Demographics and lifestyle ----------------------------------------------

demo_vars <- c("agein", "bmi", "edu3cat", "female", "black")

# Total intake and environmental impact -----------------------------------

total_vars <- c("kcal", "gram", "srv", "gw_kg", "lu_m2", "wc_m3")

# Food group variables ----------------------------------------------------

# 28 Food groups
fg_name <- grep("_gram", names(ev), value = TRUE) %>% 
  gsub("_gram", "", .)
fg_name

# Create lists of variables
kcal_vars <- paste0(fg_name, "_kcal")
gram_vars <- paste0(fg_name, "_gram")
srv_vars  <- paste0(fg_name, "_srv")
gwp_vars  <- paste0(fg_name, "_gw_kg")
lu_vars   <- paste0(fg_name, "_lu_m2")
wc_vars   <- paste0(fg_name, "_wc_m3")

# Winsorize food group variables at 99.9 percentitle
# Recalculate total kcal, gram, srv, gwp, lu, and wc
ev2 <- ev %>% 
  mutate(across(fruit_kcal:cereal_srv, Winsorize, probs = c(0, 0.999))) %>% 
  mutate(kcal = rowSums(across(all_of(kcal_vars))),
         gram = rowSums(across(all_of(gram_vars))),
         srv  = rowSums(across(all_of(srv_vars))),
         gw_kg = rowSums(across(all_of(gwp_vars))),
         lu_m2 = rowSums(across(all_of(lu_vars))),
         wc_m3 = rowSums(across(all_of(wc_vars))))

```

## Issues

Environmental variables of cereal (`cereal_gw_kg`, `cereal_lu_m2`, `cereal_wc_m3`) have all zero values. Andrew M has been notified.

## Changes
All food group variables (`*_kcal`, `*_gram`, `*_srv`, `*_gw_kg`, `*_lu_m2`, `*_wc_m3`) were winsorized at the 99.9th percentitle of each variable. Total `kcal`, `gram`, `srv`, `gw_kg`, `lu_m2`, and `wc_m3` were recalculated by summing across 28 food groups.

## Mean plots of environmental impact by food group

```{r echo = FALSE}

# Mean plots of environmental variables by food groups
MeanPlot <- function(data, vars){
  data %>% 
    select(all_of(vars)) %>% 
    summarize_all(mean) %>% 
    pivot_longer(all_of(vars), names_to = "Variable", values_to = "Mean") %>% 
    ggplot(aes(x = reorder(Variable, Mean), y = Mean)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "Food group")
}

```

* Mean GWP by food group: The consumption of `dairy` contributes to the largest GWP, followed by `beef` and `water`, among 28 food groups.

* Mean land use by food group: `dairy` followed by `veg`, `legumes` and `whlgrain`.

* Mean water consumption by food group: `fvjuice`, followed by `veg` and `fruit`.

```{r echo = FALSE, fig.height = 4, out.width=c('50%', '50%'), fig.show='hold'}
ev2 %>% MeanPlot(gwp_vars)
ev2 %>% MeanPlot(lu_vars)
ev2 %>% MeanPlot(wc_vars)
```

## Contribution of food groups (%) over total energy intake

```{r, echo = FALSE, message = FALSE}
pct_ev_plot <- function(vars, denominator, label){
  denominator <- sym(denominator)
  ylab <- paste("Food group", label, "/ Total", label, "* 100")
  tmp <- ev2 %>% 
    mutate_at(all_of(vars), ~ .x / !!denominator * 100) %>% 
    pivot_longer(all_of(vars[-28]), names_to = "Variable", values_to = "Value") %>% 
    mutate(Variable = factor(Variable, levels = vars[-28])) %>% 
    ggplot(aes(x = kcal, y = Value, color = Variable)) +
    geom_smooth() + 
    labs(x = "Total energy intake (kcal) per day", y = ylab)
  # tmp  + theme(legend.position = "bottom", legend.title = element_blank())
  tmp + facet_wrap(~Variable, ncol = 7) +
    theme(legend.position = "none")
}

```

* GWP: 

```{r echo = FALSE, message = FALSE}
pct_ev_plot(gwp_vars, "gw_kg", "GWP")
```

* Land use: 

```{r echo = FALSE, message = FALSE}
pct_ev_plot(lu_vars, "lu_m2", "LU")
```

* Water consumption: 

```{r echo = FALSE, message = FALSE}
pct_ev_plot(wc_vars, "wc_m3", "WC")
```


