---
title: 'AHS-2 Environmental Nutrition'
output: github_document
# output: pdf_document
# header-includes:
  # - \usepackage{titling}
  # - \setlength{\droptitle}{-8em} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required packages
pacs <- c("tidyverse", "readxl", "tableone", "GGally", "egg", "DescTools")
sapply(pacs, require, character.only = TRUE)

# Read data ---------------------------------------------------------------

is_local <- TRUE
zipfile  <- "./data/baseline-environmental-data-per-subject-20210912.zip"
fname    <- "baseline-environmental-data-per-subject-20210912.csv"

source("dataprep.R")

```

## Dataset
* File path: `M:\Groups\Nutrition\Environmental Nutrition\AHS-2 Environment and Health`
* File Name: ``r fname``

* Includes *n* = `r nrow(ev)` subjects and
* 190 variables:
  * `analysisid`
  * Demographics:
    * Age at baseline: `agein`
    * BMI: `bmi`
    * Education, 3 levels: `edu3cat`
    * Gender: `female`
    * Race (Black/Non-Black): `black`
  * Total intake in kcal, gram and servings per day
  * 28 food groups in:
    * kcal/day: `*_kcal`
    * gram/day: `*_gram`
    * standard servings/day: `*_srv`
    * GWP (kg CO<sub>2</sub>-eq): `*_gw_kg`
    * land use (m²a): `*_lu_m2`
    * water consumption (m³): `*_wc_m3`
    * (replace * with food group name -- see below)
    
* There are 28 food groups:

```{r echo = FALSE}
fg_name
```

### Issues

* Add an exclsion criterion on kcal after winzorization?
* Change units of environmental variables

### Changes

* More subjects (+12,000) included.
* Dietary pattern (5 groups) added.
* `pork` and `beef` intakes are now separated.
* All food group variables (`*_kcal`, `*_gram`, `*_srv`, `*_gw_kg`, `*_lu_m2`, `*_wc_m3`) were winsorized at the 99.9th percentile of each variable. Total `kcal`, `gram`, `srv`, `gw_kg`, `lu_m2`, and `wc_m3` were recalculated by summing across 28 food groups after winsorization.
* Environmental impact variables (`*_gw_kg`, `*_lu_m2`, `*_wc_m3`) were standardized to 2000 kcal/day. This was done as follows: non-zero values of each environmental variable were log-transformed (due to skewness) and then regressed on the total energy. Residuals from the regression were added by predicted values for 2000 kcal and then back-transformed to the original unit. Zero values remained zero in this process.

## Demographics

```{r echo = FALSE}
# Demographics table, unstratified
ev %>% 
  CreateTableOne(demo_vars, data = .) %>% 
  print(showAllLevels = TRUE) %>% 
  kableone()
```

## Total food intake
* Distributions of total intake in kcal, gram and servings per day
  * In the original data, those with kcal <500 or >4500 were already excluded. The total intake shown below was re-calculated by summing up all winsorized food group intakes.
  * The max gram intake became more reasonable after winsorizing data.

```{r echo = FALSE}
# Distribution of total intake
ev %>% 
  select(kcal, gram, srv) %>% 
  psych::describe(quant=c(.25,.75)) %>% 
  as.data.frame() %>% 
  select(min, Q0.25, median, Q0.75, max, mean, sd, skew) %>% 
  mutate_all(round, 2) %>% 
  knitr::kable()
```

## Total environmental impact
* Distributions of total GWP, land use and water consumption are right-skewed:

```{r echo = FALSE, fig.height = 2.3}
# Distribution of total env impact
ev %>% 
  select(gw_kg, lu_m2, wc_m3) %>% 
  psych::describe(quant=c(.25,.75)) %>%
  as.data.frame() %>% 
  select(min, Q0.25, median, Q0.75, max, mean, sd, skew) %>% 
  mutate_all(round, 2) %>% 
  knitr::kable()

ev %>% 
  select(gw_kg, lu_m2, wc_m3) %>% 
  pivot_longer(gw_kg:wc_m3, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = c("gw_kg", "lu_m2", "wc_m3"))) %>% 
  # mutate_if(is.numeric, log) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ variable, scales = "free")

```

## Distribution of environmental impact by food group

* GWP

```{r echo = FALSE}
# Descriptive stats
all_desc <- function(data, vars, digits = 2){
  data %>% 
    select(all_of(vars)) %>% 
    psych::describe(quant=c(.25,.75)) %>%
    as.data.frame() %>% 
    select(min, Q0.25, median, Q0.75, max, mean, sd, skew) %>%
    mutate_all(round, digits)
}
ev %>% all_desc(gwp_vars, digits = 3)

```

* Land use

```{r echo = FALSE}
ev %>% all_desc(lu_vars, digits = 3)
```

* Water consumption

```{r echo = FALSE}
ev %>% all_desc(wc_vars, digits = 3)
```

## Mean plots of environmental impact by food group

```{r echo = FALSE}

# Mean plots of environmental variables by food groups
MeanPlot <- function(data, vars){
  data %>% 
    select(all_of(vars)) %>% 
    summarize_all(mean) %>% 
    pivot_longer(all_of(vars), names_to = "Variable", values_to = "Mean") %>% 
    ggplot(aes(x = reorder(Variable, Mean), y = Mean)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "Food group")
}

```

* Mean GWP by food group: The consumption of `dairy` contributes to the largest GWP, followed by `whlgrain` and `beef`, among 28 food groups.

* Mean land use by food group: `dairy` followed by `veg`, `legumes` and `whlgrain`.

* Mean water consumption by food group: `fvjuice`, followed by `veg` and `fruit`.

```{r echo = FALSE, fig.height = 5, out.width=c('50%', '50%'), fig.show='hold'}
ev %>% MeanPlot(gwp_vars)
ev %>% MeanPlot(lu_vars)
ev %>% MeanPlot(wc_vars)
```

## Comparisons of environmental impact by dietary pattern

* Total energy-adjusted GWP (`gw_kg_std`), land use (`lu_m2_std`), and water consumption (`wc_m3_std`) by dietary pattern are shown in violin plots below. Note that the horizontal axis is on the pseudo-log scale.
* GWP and land use increase in the order of vegan < lacto-ovo < pesco < semi < non-veg.
  * The average total GWP of non-vegetarians is almost twice as high as that of vegans.
* For water consumption, vegans have the highest usage: vegan > pesco > lacto-ovo > semi > non-veg.

```{r echo = FALSE, fig.height = 5, fig.width = 10, message = FALSE}
ev_vars_std <- c("gw_kg_std", "lu_m2_std", "wc_m3_std")
ev %>% 
  pivot_longer(ev_vars_std, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = ev_vars_std)) %>% 
  filter(!is.na(vegstat)) %>% 
  ggplot(aes(x = vegstat, y = value, fill = vegstat)) +
  geom_violin() +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  coord_flip() +
  scale_x_discrete(limits = rev(levels(ev$vegstat))) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10)) +
  facet_grid(~variable, scales = "free") +
  labs(x = "Dietary pattern") +
  theme(legend.position = "none")

ev %>% 
  pivot_longer(ev_vars_std, names_to = "variable", values_to = "value") %>% 
  mutate(variable = factor(variable, levels = ev_vars_std)) %>% 
  filter(!is.na(vegstat)) %>% 
  group_by(variable, vegstat) %>% 
  summarize(Median = median(value), Mean = mean(value), SD = sd(value)) %>% 
  mutate_if(is.numeric, round, 2)
```

* In all three variables, differences across dietary patterns were highly significant:

```{r echo = FALSE}
# Kruskal-Wallis to compare across vegstat
ev[ev_vars_std] %>% 
  map_dfr(\(x) broom::tidy(kruskal.test(x ~ vegstat, data = ev))) %>% 
  mutate(Variable = ev_vars_std, p.value = Hmisc::format.pval(p.value)) %>% 
  select(Variable, method, statistic, p.value) %>% 
  knitr::kable()
```

## Correlation heatmap of food group intake

* Correlation heat maps were generated among 28 food groups in terms of gram (standardized to 2000 kcal), based on Spearman correlation. The order of food group variables were determined by the average-linkage hierarchical clustering.

* Moderately high correlations were found between beef and poultry (*r* = 0.66), between poultry and fish (*r* = 0.60), and between beef and processed meat (*r* = 0.57). Notice the groupings of positively correlated foods (3 groupings were chosen from the cluster analysis).

```{r echo = FALSE, message = FALSE}
ev_heatmap <- function(data, vars, gsubstr, addrect){
  data %>% 
  select(all_of(vars)) %>% 
  rename_with(~gsub(gsubstr, "", .x)) %>% 
  cor(method = "spearman") %>% 
  corrplot::corrplot(method = "color", order = "hclust", hclust.method = "average", addrect = addrect, tl.col = "black", tl.cex = 0.8)
}

ev %>% ev_heatmap(gram_vars_std, "_gram_std", addrect = 3)
```

## Linear models for env variables on food group intakes

### Models adjusting for the total energy intake

* A linear model was fitted for GWP, regressed on all 28 food group intakes (standardized to 2000 kcal). The unit of GWP was changed to g CO<sub>2</sub>-eq  to make beta coefficients larger. Estimated beta coefficients (and their 95% CI) were plotted from the largest to the smallest.

```{r echo = FALSE, message = FALSE}
# Model setup
dv    <- "gw_kg_std"
covar <- c("kcal")
rhs <- paste(c(gram_vars_std, covar), collapse = " + ")
fm1 <- formula(paste(dv, "~", rhs))

# Change units before lm fit...
ev_lm <- ev %>% 
  mutate(kcal = kcal / 1000,
         gw_kg_std = gw_kg_std * 1000)

# Function to tidy beta estimates
extract_beta <- function(model){
  broom::tidy(model, conf.int = TRUE) %>% 
    slice(-1, -n()) %>% 
    select(term, estimate, conf.low, conf.high) %>%
    mutate(term = gsub("_gram_std", "", term)) %>% 
    arrange(-estimate) %>% 
    mutate(term = factor(term))
}

# Initial model
gw_mod1 <- lm(fm1, data = ev_lm) 
gw_mod1_coef <- extract_beta(gw_mod1) 

# log transformation? -- not used here
gw_mod2 <- update(gw_mod1, log(.) ~ ., data = ev_lm) 
gw_mod2_coef <- extract_beta(gw_mod2) 

# Plot beta estimates
gw_mod1_coef %>% 
  ggplot(aes(x = forcats::fct_reorder(term, estimate), y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) + 
  coord_flip() +
  labs(x = "Intake of food groups (gram/day)",
       y = "Beta coefficients for GWP (with 95% CI)")
```

```{r echo = FALSE}
gw_mod1_coef %>% 
  mutate_if(is.numeric, round, 2) %>% 
  knitr::kable()
```

### Models without kcal

 * Estimated beta coefficients were compared between the models with and without the total energy intake. Generally speaking, further adjusting for kcal appears to attenuate the beta estimates slightly.

```{r echo = FALSE, message = FALSE}
# Remove kcal from the model
gw_mod3 <- update(gw_mod1, . ~ . -kcal, data = ev_lm) 
gw_mod3_coef <- extract_beta(gw_mod3) 

# Compare coefficient estimates
gw_mod1_coef %>% 
  mutate(model = "with Kcal") %>% 
  bind_rows(gw_mod3_coef %>% mutate(model = "without kcal")) %>% 
  ggplot(aes(x = forcats::fct_reorder(term, estimate), y = estimate, color = model)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) + 
  coord_flip() +
  labs(x = "Intake of food groups (gram/day)",
       y = "Beta coefficients for GWP (with 95% CI)")
```

### Models excluding pork eaters

 * Because of the apparent negative association between GWP and pork intake, the model was refit after removing any pork eaters (n = 2313 or 2.6%). Estimated beta coefficients were virtually unchanged after the exclusion.

```{r echo = FALSE, message = FALSE}
# Remove pork from the model
gw_mod4 <- update(gw_mod1, . ~ . -pork_gram_std, data = ev_lm[ev_lm$pork_gram == 0,]) 
gw_mod4_coef <- extract_beta(gw_mod4) 

# Compare coefficient estimates
gw_mod1_coef %>% 
  mutate(model = "Include pork eaters") %>% 
  bind_rows(gw_mod4_coef %>% mutate(model = "Exclude pork eaters")) %>% 
  ggplot(aes(x = forcats::fct_reorder(term, estimate), y = estimate, color = model)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) + 
  coord_flip() +
  labs(x = "Intake of food groups (gram/day)",
       y = "Beta coefficients for GWP (with 95% CI)")
```


## GAMLSS models on GWP

* The model fit was poor, as expected from the distribution of GWP. Model assumptions were clearly violated.

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 10}
ggResidpanel::resid_panel(gw_mod1, plots = c("resid", "qq"))
```

* Instead of assuming normal errors, we used:
  * the generalized gamma (GG) model
  * the generalized beta of the second kind (GB2) model

* A worm plot ([van Buuren & Fredriks, 2001](https://pubmed.ncbi.nlm.nih.gov/11304741/)) was generated for each model to compare model fits.
  * Compared to the normal model, the GG model fits better, having a smaller AIC. However, its worm plot still indicates a poor fit.
  * The model fit further improved in the GB2 model. Its worm plot is more acceptable, compared to the other two models.

```{r echo = FALSE, message = FALSE, warning = FALSE, results = FALSE, error = TRUE, fig.height = 5, fig.width = 5, out.width=c('50%', '50%'), fig.show='hold'}
library(gamlss)

# Data for gamlss models
ev_gamlss <- ev %>% 
  dplyr::select(all_of(dv), all_of(covar), all_of(gram_vars_std)) %>% 
  mutate(kcal = kcal / 1000,
         gw_kg_std = gw_kg_std * 1000)

# Normal error
gw_mod1 <- gamlss(fm1, family = NO, data = na.omit(ev_gamlss)) 

# Generalized Gamma distribution (three parameters)
gw_mod2 <- gamlss(fm1, family = GG, data = na.omit(ev_gamlss))

# Generalized beta of the second kind (four parameters)
gw_mod3 <- gamlss(fm1, family = GB2, method = mixed(5, 30), data = na.omit(ev_gamlss))

wp(gw_mod1, xlim.all = 5, ylim.all = 7); title(paste("Normal: AIC = ", round(gw_mod1$aic)))
wp(gw_mod2, xlim.all = 5, ylim.all = 7); title(paste("GG: AIC = ", round(gw_mod2$aic)))
wp(gw_mod3, xlim.all = 5, ylim.all = 2); title(paste("GB2: AIC = ", round(gw_mod3$aic)))
```
